Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
small mods.

If we assume an algorithm where the rows are processed the one after
the other, the amount of memory written is precisely the
size of the {\tt y} vector. (This assumption removes the possibility of
blocking or cache partitioning techniques.) $$b_{wT} = s_y = n_v n_m b_x n_r$$

@@

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of nonzeros in the
matrix. Note however that most architectures cannot read memory a
single byte at a time. Instead, a minimum number of bytes, equal to
the size of a cacheline $cl$, are transfered at once.  When there are
multiple vectors, each nonzero element uses $n_v$ consecutive
entries. The worst case number of bytes read and transfered are
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

@@

Figure~\ref{fig:ratio_bytes_flops} presents the flop to byte ratios
for single precision computation in the best and the worst case on a
classical cache based architecture ($cl = 64$) and assuming $cl=1$. We
can easily see the potential improvement in the computational
intensity when the number of matrices of vectors increase. There is a
significant difference between the best and the worst case: there is a
$8$ fold difference in the 1 matrix 1 vector case but that gap closes
with the increase in the number of vectors and matrices to a 4 fold
difference in the 4 matrices and 4 vectors case and 1.6 fold in the 16
matrices and 16 vectors. The ratios computed with $cl = 1$ are mostly
similar to the the one with $cl=64$ when the number of vectors is
large. But the differences are important when the number of vector is
low: this highlight that accessing the memory per batch is the main
problem faced when performing a standard SpMV computation.

@@

  \caption{Ratio of flops to bytes in single precision: in the best case in
    \ref{fig:ratio_best} best; in the worst case when $cl=64$ in
    \ref{fig:ratio_worst}; and in the worst case neglecting that the
    memory transfers are by batch (equivalent to $cl = 1$) in
    \ref{fig:ratio_worst_nocache}.}
  \label{fig:ratio_bytes_flops}
\end{figure*}

Figure~\ref{fig:ratio_bytes_flops} shows how these ratios translate
into actual performance. This figure provides projected best and worst
Gflop/s achievable assuming the computation is memory bound and the
architecture reaches 150Gflop/s. Notice that \cite{Saule13-ARXIV}
showed a higher peak bandwidth, but we will show in
Section~\ref{sec:expe} why 150Gflop/s is a better estimate of what one
might achieve in these kernels. In single precision, the worse that
one could achieve by using 4 vectors and 4 matrices is much higher
than the best achievable using a classical SpMV computation. The best
performance reachable is 210Gflop/s: almost 6 times higher than the
peak of the classical SpMV case.

>>>>

If we assume an algorithm where the rows are processed one after
the other, the amount of memory written is precisely the
size of the {\tt y} vector. (This assumption removes the possibility of
blocking or cache partitioning techniques.) $$b_{wT} = s_y = n_v n_m b_x n_r$$

@@

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of nonzeros in the
matrix. Note however that most architectures cannot read memory a
single byte at a time. Instead, a minimum number of bytes, equal to
the size of a cacheline $cl$, are transfered at once.  When there are
multiple vectors, each nonzero element uses $n_v$ consecutive
entries. In the worst case, the number of bytes read and transfered is
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

@@

Figure~\ref{fig:ratio_bytes_flops} presents the flop to byte ratios
for single precision computation in the best and the worst case on a
classical cache based architecture ($cl = 64$) and assuming $cl=1$. We
can easily see the potential improvement in the computational
intensity when the number of matrices of vectors increase. There is a
significant difference between the best and the worst case: there is a
eightfold difference in the one matrix -- one vector (1/1) case but that gap closes
with the increase in the number of vectors and matrices to a fourfold
difference in the four4matrices and four vectors (4/4) case and 1.6 fold in the 
16/16 case. 
The ratios computed with $cl = 1$ are mostly
similar to the one with $cl=64$ when the number of vectors is
large. But the differences are important when the number of vectors is
low: this highlights that accessing the memory per batch \todo{what do you mean by 
accessing the memory per batch} is the main
problem faced when performing a standard SpMV computation.

@@

  \caption{Ratio of flops to bytes in single precision:  
    \ref{fig:ratio_best} best case; \ref{fig:ratio_worst} worst case when $cl=64$ 
    ; and \ref{fig:ratio_worst_nocache} worst case neglecting that the
    memory transfers are by batch (equivalent to $cl = 1$)}
  \label{fig:ratio_bytes_flops}
\end{figure*}

Figure~\ref{fig:ratio_bytes_flops} shows how these ratios translate
into actual performance. This figure provides projected best and worst
Gflop/s achievable assuming the computation is memory-bound and the
architecture reaches 150 Gflop/s. Notice that \cite{Saule13-ARXIV}
showed a higher peak bandwidth, but we will show in
Section~\ref{sec:expe} why 150Gflop/s is a better estimate of what one
might achieve in these kernels. In single precision, the worse that
one could achieve by using 4 vectors and 4 matrices is much higher
than the best achievable using a classical SpMV computation. The best
performance reachable is 210Gflop/s: almost six times higher than the
peak of the classical SpMV case.
